\chapter{Clocks}\label{XtenClocks}\index{clock}

Many concurrent algorithms proceed in phases: in phase {$k$}, several
activities work independently, but synchronize together before proceeding on
to phase {$k+1$}. X10 supports this communication structure (and many
variations on it) with a generalization of barriers 
called {\em clocks}. Clocks are designed so that programs which follow a
simple syntactic discipline will not have either deadlocks or race conditions.


The following minimalist example of clocked code has two worker activities A
and B, and three phases. In the first phase, each worker activity says its
name followed by 1; in the second phase, by a 2, and in the third, by a 3.  
So, if \xcd`say` prints its argument, 
\xcd`A-1 B-1 A-2 B-2 B-3 A-3`
would be a legitimate run of the program, but
\xcd`A-1 A-2 B-1 B-2 A-3 B-3`
(with \xcd`A-2` before \xcd`B-1`) would not.

The program creates a clock \xcd`cl` to manage the phases.  Each participating
activity does
the work of its first phase, and then executes \xcd`Clock.advanceAll();` to
signal that it 
is finished with that work. \xcd`Clock.advanceAll();` is blocking, and causes the participant to
wait until all participant have finished with the phase -- as measured by the
clock \xcd`cl` to which they are both registered.  
Then they do the second phase, and another \xcd`Clock.advanceAll();` to make sure that
neither proceeds to the third phase until both are ready.  This example uses
\xcd`finish` to wait for both particiants to finish.  




%~~gen ^^^ Clocks10
%package Clocks.For.Spock;
%~~vis
\begin{xten}
class ClockEx {
  static def say(s:String) = 
     { atomic{x10.io.Console.OUT.println(s);} }
  public static def main(argv:Rail[String]) {
    finish async{
      val cl = Clock.make();
      async clocked(cl) {// Activity A
        say("A-1");
        Clock.advanceAll();
        say("A-2");
        Clock.advanceAll();
        say("A-3"); 
      }// Activity A

      async clocked(cl) {// Activity B
        say("B-1");
        Clock.advanceAll();
        say("B-2");
        Clock.advanceAll();
        say("B-3"); 
      }// Activity B
    }
  }
 }
\end{xten}
%~~siv
%~~neg

This chapter describes the syntax and semantics of clocks and
statements in the language that have parameters of type \xcd"Clock". 

The key invariants associated with clocks are as follows.  At any
stage of the computation, a clock has zero or more {\em registered}
activities. An activity may perform operations only on those clocks it
is registered with (these clocks constitute its {\em clock set}). 
An attempt by an activity to operate on a clock it is not registered with
will cause a 
\xcd"ClockUseException"\index{clock!ClockUseException}
to be thrown.  
An activity is registered with zero or more clocks when it is created.
During its lifetime the only additional clocks it can possibly be registered with
are exactly those that it creates. In particular it is not possible
for an activity to register itself with a clock it discovers by
reading a data structure.

The primary operations that an activity \xcd`a` may perform on a clock \xcd`c`
that it is registered upon are: 
\begin{itemize}
\item It may spawn and simultaneously  {\em register} a new activity on
      \xcd`c`, with the statement       \xcd`async clocked(c)S`.
\item It may {\em unregister} itself from \xcd`c`, with \xcd`c.drop()`.  After
      doing so, it can no longer use most primary operations on \xcd`c`.
\item It may {\em resume} the clock, with \xcd`c.resume()`, indicating that it
      has finished with the current phase associated with \xcd`c` and is ready
      to move on to the next one.
\item It may {\em wait} on the clock, with \xcd`c.advance()`.  This first does
      \xcd`c.resume()`, and then blocks the current activity until the start
      of the next phase, \viz, until all other activities registered on that
      clock have called \xcd`c.resume()`.
\item It may {\em block} on all the clocks it is registered with
      simultaneously, by the command \xcd`Clock.advanceAll();`.  This, in effect, calls
      \xcd`c.advance()` simultaneously 
      on all clocks \xcd`c` that the current activity is registered with.
\item Other miscellaneous operations are available as well; see the
      \xcd`Clock` API.
\end{itemize}


\section{Clock operations}\label{sec:clock}
\index{clock!operations on}
There are two language constructs for working with clocks. 
\xcd`async clocked(cl) S` starts a new activity registered on one or more
clocks.  \xcd`Clock.advanceAll();` blocks the current activity until all the activities
sharing clocks with it are ready to proceed to the next clock phase. 
Clocks are objects, and have a number of useful methods on them as well.

\subsection{Creating new clocks}\index{clock!creation}\label{sec:clock:create}

Clocks are created using a factory method on \xcd"x10.lang.Clock":


%~~gen ^^^ Clocks20
% package Clocks.For.Spocks;
%class Clockuser {
% def example() {
%~~vis
\begin{xten}
val c: Clock = Clock.make();
\end{xten}
%~~siv
%}}
%~~neg


The current activity is automatically registered with the newly
created clock.  It may deregister using the \xcd"drop" method on
clocks (see the documentation of \xcd"x10.lang.Clock"). All activities
are automatically deregistered from all clocks they are registered
with on termination (normal or abrupt).

\subsection{Registering new activities on clocks}
\index{clock!clocked statements}\label{sec:clock:register}

%##(AsyncStatement ClockedClause
\begin{bbgrammar}
%(FROM #(prod:AsyncStmt)#)
           AsyncStmt \: \xcd"async" ClockedClause\opt Stmt & (\ref{prod:AsyncStmt}) \\
                     \| \xcd"clocked" \xcd"async" Stmt \\
%(FROM #(prod:ClockedClause)#)
       ClockedClause \: \xcd"clocked" Arguments & (\ref{prod:ClockedClause}) \\
\end{bbgrammar}
%##)

The \xcd`async` statement with a \xcd`clocked` clause of either form, say 
%~~gen ^^^ Clocks30
%package Clocks.For.Jocks;
%class Qlocked{
%static def S():void{}
%static def flock() { 
% val c1 = Clock.make(), c2 = Clock.make(), c3 = Clock.make();
%~~vis
\begin{xten}
  async clocked (c1, c2, c3) S
\end{xten}
%~~siv
%();
%}}
%~~neg
starts a new activity, initially registered with
clocks \xcd`c1`, \xcd`c2`, and \xcd`c3`, and  running \xcd`S`. The activity running this code must
be registered on those clocks. 
Violations of these conditions are punished by the throwing of a
\xcd"ClockUseException"\index{clock!ClockUseException}. 

% An activity may transmit only those clocks that are registered with and
% has not quiesced on (\Sref{resume}). 
% A \xcd"ClockUseException"\index{clock!ClockUseException} is
%thrown if (and when) this condition is violated.

If an activity {$a$} that has executed \xcd`c.resume()` then starts a
new activity {$b$} also registered on \xcd`c` (\eg, via \Xcd{async
clocked(c) S}), the new activity {$b$} starts out having also resumed
\xcd`c`, as if it too had executed \xcd`c.resume()`. That is, $a$ and {$b$}
are in the same phase of the clock. 
%~~gen ^^^ Clocks40
% package Clocks.For.Jocks.In.Clicky.Smocks;
%class Example{
%static def S():void{}
%static def a_phase_two():void{}
%static def b_phase_two():void{}
%static def example() {
%~~vis
\begin{xten}
// ACTIVITY a
val c = Clock.make();
c.resume();
async clocked(c) {
  // ACTIVITY b
  c.advance();
  b_phase_two();
  // END OF ACTIVITY b
}
c.advance();
a_phase_two();
// END OF ACTIVITY a
\end{xten}
%~~siv
%} }
%~~neg
In the proper execution, {$a$} and {$b$} both perform
\xcd`c.advance()` and then their phase-2 actions.  
However, if {$b$} were not
initially in the resume state for \xcd`c`, there would be a race condition;
{$b$} could perform \xcd`c.advance()` and proceed to \xcd`b_phase_two`
before {$a$} performed \xcd`c.advance()`.


An activity may check whether or not it is registered on a clock \xcd"c" by
%~~exp~~`~~`~~c:Clock ~~ ^^^ Clocks50
the method call \xcd`c.registered()`


\begin{note}
\Xten{} does not contain a ``register'' operation that would allow an activity
to discover a clock in a datastructure and register itself (or another
process) on it. Therefore,
while a clock \xcd`c` may be stored in a data structure by one activity
\xcd`a` and read from it by another activity \xcd`b`, \xcd`b` cannot do much
with \xcd`c` unless it is already registered with it.  In particular, it
cannot register itself on \xcd`c`, and, lacking that registration, cannot
register a sub-activity on it with \xcd`async clocked(c) S`.
\end{note}


\subsection{Resuming clocks}\index{clock!resume}\label{resume}\label{sec:clock:resume}
\Xten{} permits {\em split phase} clocks. An activity may wish
to indicate that it has completed whatever work it wishes to perform
in the current phase of a  clock \xcd"c" it is registered with, without
suspending itself altogether. It may do so  by executing 
%~~stmt~~`~~`~~c:Clock ~~ ^^^ Clocks60
\xcd`c.resume();`.



An activity may invoke \xcd`resume()` only on a clock it is registered with,
and has not yet dropped (\Sref{sec:clock:drop}). A
\xcd"ClockUseException"\index{clock!ClockUseException} is thrown if this
condition is violated. Nothing happens if the activity has already invoked a
\xcd"resume" on this clock in the current phase.

\subsection{Advancing clocks}\index{clock!advanceAll}\label{sec:clock:next}
An activity may execute the following method call to signal that it is done
with the current phase.
\begin{xten}
Clock.advanceAll();
\end{xten}

\noindent 
Execution of this call blocks until all the clocks that the
activity is registered with (if any) have advanced. (The activity
implicitly issues a \xcd"resume" on all clocks it is registered
with before suspending.)

\xcd`Clock.advanceAll();` may be thought of as calling \xcd`c.advance()` in parallel for all
clocks that the current activity is registered with.  (The parallelism is
conceptually important: if activities {$a$} and {$b$} are both
registered on clocks \xcd`c` and \xcd`d`, and {$a$} executes
\xcd`c.advance(); d.advance()` while {$b$} executes \xcd`d.advance(); c.advance()`,
then the two will deadlock.  However, if the two clocks are waited on in
parallel, as \xcd`Clock.advanceAll();` does, {$a$} and {$b$} will not deadlock.)

Equivalently, \xcd`Clock.advanceAll();` sequentially calls \xcd`c.resume()` for each
registered clock \xcd`c`, in arbitrary order, and then \xcd`c.wait()` for each
clock, again in arbitrary order.  



An activity blocked on \xcd"advance()" resumes execution once
it is marked for progress by all the clocks it is registered with.

\subsection{Dropping clocks}\index{clock!drop}\label{sec:clock:drop}
%~~stmt~~`~~`~~ c:Clock~~ ^^^ Clocks70
An activity may drop a clock by executing \xcd`c.drop();`.



\noindent{} The activity is no longer considered registered with this
clock.  A \xcd"ClockUseException" is thrown if the activity has
already dropped \xcd"c".

\section{Deadlock Freedom}

In general, programs using clocks can deadlock, just as programs using loops
can fail to terminate.  However, programs written with a particular syntactic
discipline {\em are} guaranteed to be deadlock-free, just as programs which
use only bounded loops are guaranteed to terminate.  The syntactic discipline
is: 
\begin{itemize}
\item The \xcd`advance()` {instance method} shall not be called on any clock.
      (The 
      \xcd`Clock.advanceAll();` method is allowed for this discipline.)
\item Inside of \xcd`finish{S}`, all clocked \xcd`async`s shall be in the scope
      an unclocked \xcd`async`.  
\end{itemize}
X10 does not enforce this discipline.  Doing so would exclude useful programs,
many of which are deadlock-free for reasons more subtle than the
straightforward syntactic discipline.  Still, this discipline is useful for
simple cases. 

The first clause of the discipline prevents a deadlock in which an activity
is registered on two clocks, advances one of them, and ignores the other. 
The second clause prevents the following deadlock.  
%~~gen ^^^ Clocks80
% package Clocks.Finish.Hates.Clocks;
% class Example{
% def example() {
%~~vis
\begin{xten}
val c:Clock = Clock.make();
async clocked(c) {                // (A) 
      finish async clocked(c) {   // (B) Violates clause 2
            Clock.advanceAll();   // (Bnext)
      }
      Clock.advanceAll();         // (Anext)
}
\end{xten}
%~~siv
% } } 
%~~neg
\xcd`(A)`, first of all, waits for the \xcd`finish` containing \xcd`(B)` to
finish.  
\xcd`(B)` will execute its \xcd`advance` at \xcd`(Bnext)`, and then wait for all
other activities registered on \xcd`c` to execute their \xcd`advance()`s.
However, \xcd`(A)` is registered on \xcd`c`.  So, \xcd`(B)` cannot finish
until \xcd`(A)` has proceeded to \xcd`(Anext)`, and \xcd`(A)` cannot proceed
until \xcd`(B)` finishes. Thus, this causes deadlock.


\section{Program equivalences}
From the discussion above it should be clear that the following
equivalences hold:

{\small
\begin{eqnarray}
 \mbox{\xcd"c.resume(); Clock.advanceAll();"}       &=& \mbox{\xcd"Clock.advanceAll();"}\\
 \mbox{\xcd"c.resume(); d.resume();"} &=& \mbox{\xcd"d.resume(); c.resume();"}\\
 \mbox{\xcd"c.resume(); c.resume();"} &=& \mbox{\xcd"c.resume();"}
\end{eqnarray}
}

Note that \xcd"Clock.advanceAll(); Clock.advanceAll();" is not the same as \xcd"Clock.advanceAll();". The
first will wait for clocks to advance twice, and the second
once.  

%\notinfouro{\input{clock/imp-notes.tex}}
%\notinfouro{\input{clock/clocked-types.tex}}
%\notinfouro{\input{clock/examples.tex}}

\section{Clocked Finish}
\index{finish!clocked}
\index{async!clocked}
\index{clocked!finish}
\index{clocked!async}
\label{ClockedFinish}

In the most common case of a single clock coordinating a few behaviors, X10
allows coding with an implicit clock.  \xcd`finish` and \xcd`async` statements
may be qualified with \xcd`clocked`.  

A \xcd`clocked finish` introduces a new clock.  It executes its body in the
usual way that a \xcd`finish` does--- except that, when its body completes,
the activity executing the \xcd`clocked finish` drops the clock, while it
waits for asynchronous spawned \xcd`async`s to terminate.  

A \xcd`clocked async` registers its async with the implicit clock of
the surrounding \xcd`clocked finish`.   

The bodies of the \xcd`clocked finish` and \xcd`clocked async`
statements  may use the
\xcd`Clock.advanceAll()` 
method call to advance the implicit clock.  Since the implicit clock is not
available in a variable, it cannot be manipulated directly. (If you want to
manipulate the clock directly, use an explicit clock, not a \xcd`clocked finish`.)

\begin{ex}
The following code starts two activities, each of which perform their first
phase, wait for the other to finish phase 1, and then perform their second
phase.  
%~~gen ^^^ Clocks90
%package Clocks.ClockedFinish;
%class Example{
%static def phase(String, Int) {}
%def example() {
%~~vis
\begin{xten}
clocked finish {
  clocked async {
     phase("A", 1);
     Clock.advanceAll();
     phase("A", 2);
  }
  clocked async {
     phase("B", 1);
     Clock.advanceAll();
     phase("B", 2);
  }
}
\end{xten}
%~~siv
%}}
%~~neg
\end{ex}

\index{finish!nested clocked}
\index{clocked finish!nested}

Clocked finishes may be nested.  The inner \xcd`clocked finish` operates in a
single phase of the outer one.  
