\chapter{Changes}

This document summarizes the main changes between X10 2.0.6 and X10 2.1.  The
descriptions are intended to be suggestive rather than definitive; see the
language specification -- when it is finished -- for full details.

\section{Object Model}

\begin{enumerate}
\item Objects are now local rather than global.
   
    \begin{enumerate}
    \item The \Xcd{home} property is gone.
    \item \Xcd{at(P)S} produces deep copies of all objects located \Xcd{here}
          when it executes \Xcd{S}.  ({\bf Warning:} They are copied even in 
          \Xcd{at(here)S}.)
    \end{enumerate}

\item The \Xcd{GlobalRef[T]} struct is the only way to produce or manipulate
      cross-place references.
    \begin{enumerate}
    \item \Xcd{GlobalRef}'s have a \Xcd{home} property.
    \item Use \Xcd{GlobalRef[Foo](foo)} to make a new global reference.
    \item Use \Xcd{myGlobalRef()} to access the object referenced; this
          requires \Xcd{here == myGlobalRef.home}. 
    \end{enumerate}


\item All those cursed \Xcd{!}s in types are gone.  

\item \Xcd{global} modifiers are now gone:
    
    \begin{enumerate}
    \item \Xcd{global} methods in {\em interfaces} are now the default. 
    \item \Xcd{global} {\em fields} are gone.  In some cases object copying
          will produce the same effect as global fields.  In other cases code
          must be rewritten.  It may be desirable to mark nonglobal fields
          \Xcd{transient} in many cases.
    \item \Xcd{global} {\em methods} are now marked \Xcd{@Global} instead.  
          Methods intended to be non-global may be marked \Xcd{@Pinned}.
    \end{enumerate}


\end{enumerate}

\section{Constructors}


\begin{enumerate}
\item \Xcd{proto} types are gone.
\item Constructors and the methods they call must satisfy a number of static
      checks.  
    
    \begin{enumerate}
    \item Constructors can only invoke \Xcd{private} or \Xcd{final} methods, 
          or methods annotated \xcd`@NonEscaping("v1,v2")`.  
    \item Methods invoked by constructors cannot read fields before they are
          written. 
    \item The compiler ensures this with a detailed protocol. 
    \end{enumerate}

\item It is still impossible for X10 constructors to leak references to
      \Xcd{this} or observe uninitialized fields of an object.  Now, however,
      the mechanisms enforcing this are less obtrusive than in 2.0.6; the
      burden is largely on the compiler, not the programmer.
\end{enumerate}




\section{Call by Reference}

A very limited form of call-by-reference is now available.


\begin{enumerate}

\item Formal parameters to functions and methods may be \Xcd{ref} rather than
      \Xcd{var} or \Xcd{val}.  
\item Assignment to a \Xcd{ref} parameter \Xcd{x} changes the original
      location that the \Xcd{ref} refers to.  \eg, 
      \xcd`def inc(ref x:Int) { x ++; }`
      allows a call \Xcd{inc(n)} to increment a local \Xcd{var} \Xcd{n}.
\item Only local variables or \Xcd{ref} parameters can be passed as actual
      \Xcd{ref} parameters.  Fields, array elements, and other variable-like
      items cannot be. 
\item External \Xcd{ref} variables cannot be captured in closures. However,
      closures may have \Xcd{ref} parameters.
\item \Xcd{ref}s are {\em not} first-class objects in X10. They cannot be
      returned from functions, stored in data structures, etc.
\item These restrictions limit the possibilities of aliasing and the need for
      boxing of \Xcd{ref} parameters.  \Xcd{ref}s to stack locations cannot,
      with these restrictions, live past the death of the location's
      containing stack frame.      
\item This allows the implementation of many core constructs as syntactic
      sugar on library calls.   Programmers may use it, but mutability should
      generally be encapsulated inside objects rather than \Xcd{ref}
      parameters. 
\end{enumerate}


\section{Accumulator Variables}

Accumulator variables generalize and make explicit collecting \Xcd{finish} in
X10 2.0.6.  An \Xcd{acc} variable is declared: 
\begin{xten}
acc(r) A;
\end{xten}
where \Xcd{r} is a {\em reducer} (much as in 2.0.6): 
\begin{xten}
struct Reducer[T](zero:T, apply:global (T,T)=>T){}
\end{xten}

Usage of \Xcd{A} is restricted in ways that make it determinate in the
intended case of a pure, associative, commutative \Xcd{apply} with unit
element \Xcd{zero}.  

\begin{enumerate}
\item \Xcd{A} is initialized to \Xcd{r.zero}.  
\item Multiple activities can {\em write} into \Xcd{A}.  In particular, the
      ``assignment'' \Xcd{A = v} is approximately interpreted as 
      \xcd`atomic{A = r.apply(A, v)}` --- that is, it accumulates \Xcd{v} into
      \Xcd{A} using \Xcd{r.apply.}
\item {\em Reading} of \Xcd{A} is restricted to situations where it makes
      sense.  Specifically, only the activity in which \Xcd{A} is declared can
      read from it, and it can only do so when all asyncs which it has spawned
      have terminated -- \eg, outside of the scope of all \Xcd{async}s and
      \Xcd{finish}es.  
\item Formal parameters of functions may be marked \Xcd{acc x:T}.  The reducer
      \Xcd{r} must not be specified; it is passed as an implicit parameter
      going with the actual \Xcd{acc} variable.  
\item X10 provides protocols for indexed collections of \Xcd{acc} variables,
      presented as objects.
\end{enumerate}



\section{Implicit clocks for each finish}


Clocks are no longer explicit objects.  Many clock operations are available
using implicit clocks.

\begin{enumerate}
\item A \Xcd{finish} may be qualified with \Xcd{clocked}, which gives it a
      clock.
\item An \Xcd{async} in a \Xcd{clocked finish} may be marked \Xcd{clocked}.
      This registers it on the same clock as the enclosing \Xcd{finish}.  
\item \xcd`clocked async S` and \xcd`clocked finish S` may use \xcd`next` in
      the body of \Xcd{S} to advance the clock.
\item When the body of a \Xcd{clocked finish} completes, the \Xcd{clocked
      finish} is dropped form the clock.  It will still wait for spawned
      asyncs to terminate, but such asyncs need to wait for it.
\end{enumerate}


\section{Clocked local variables}

Local \Xcd{val} and \Xcd{acc} variables may be \Xcd{clocked}.  They are
associated with the clock of the surrounding \Xcd{clocked finish}.  
Clocked variables have a {\em current} value and an {\em upcoming} value.  The
current value may be read at suitable times; the upcoming value may be
updated.  The \Xcd{next} phase makes the upcoming value current.

\section{Asynchronous initialization of val}

\Xcd{val}s can be initialized asynchronously.   As always with \Xcd{val}s,
they can only be read after is guaranteed that they have been initialized.
For example, both of the \Xcd{print}s below are good.  However, the
commented-out \Xcd{print} in the \Xcd{async} is bad, since it is possible that
it will be executed before the initialization of \Xcd{a}. 
\begin{xten}
val a:Int;
finish {
  async {
     a = 1; 
     print("a=" + a);
  }
  async {
     // WRONG: print("a=" + a);
  }
}
print("a=" + a);
\end{xten}



\section{Main Method}

The signature for the \Xcd{main} method is now: 
\begin{xten}
           def main(Array[String](1)) {..}
\end{xten}
or, if the arguments are actually used, 
\begin{xten}
           def main(argv: Array[String](1)) {..}
\end{xten}

\section{Removed Topics}

The following are gone: 

\begin{enumerate}
\item \Xcd{x10.lang.Clock}.
\item \Xcd{clocked} clause on \Xcd{async}, \Xcd{ateach}, and \Xcd{foreach}.
\item \Xcd{foreach} is gone.
\item All \Xcd{var}s are effectively \Xcd{shared}, so \Xcd{shared} is gone.
\item collecting \Xcd{finish}, \Xcd{offer}, and \Xcd{offers} are gone.  Use
      \Xcd{acc} variables instead.
\item The place clause on \Xcd{async} is gone.  \Xcd{async (P) S} should be
      written \Xcd{at(P) async S}.
\item Checked exceptions are gone.
\end{enumerate}


\section{Assorted Additions}


\begin{enumerate}



\item structs can have \Xcd{var} fields.  This goes with call-by-reference. 

\item Statements can be annotated \Xcd{@det}; the compiler will check them for
      determinacy. 



\end{enumerate}
