% First page

\thispagestyle{empty}

% \todo{"another" report?}

\title{ \Xten Language Specification \\
\large Version \integerversion}
%\ifdraft
\author{Vijay Saraswat, Bard Bloom, Igor Peshansky, Olivier Tardieu, and David Grove\\
\\
Please send comments to 
\texttt{vsaraswa@us.ibm.com}}
%\else
%\author{
%Vijay Saraswat \\
%Please send comments to \\
%\texttt{vsaraswa@us.ibm.com}}
%\fi
%\date{\today}

\maketitle

\newcommand\authorsc[1]{#1}
%\newcommand\authorsc[1]{\textsc{#1}}

This report provides a description of the programming
language \Xten. \Xten{} is a class-based object-oriented
programming language designed for high-performance, high-productivity
computing on high-end computers supporting $\approx 10^5$ hardware threads
and $\approx 10^{15}$ operations per second. 

\Xten{} is based on state-of-the-art object-oriented programming
languages and deviates from them only as necessary to support its
design goals. The language is intended to have a simple and clear
semantics and be readily accessible to mainstream OO programmers. It
is intended to support a wide variety of concurrent programming
idioms.
%, incuding data parallelism, task parallelism, pipelining.
%producer/consumer and divide and conquer.

%We expect to revise this document in the light of experience gained in implementing
%and using this language.

The \Xten{} design team consists of
\authorsc{David Grove},
\authorsc{Ben Herta},
\authorsc{Louis Mandel},
\authorsc{Josh Milthorpe},
\authorsc{Vijay Saraswat},
\authorsc{Avraham Shinnar},
\authorsc{Mikio Takeuchi},
\authorsc{Olivier Tardieu}.

Past members include
\authorsc{Shivali Agarwal}, 
\authorsc{Bowen Alpern}, 
\authorsc{David Bacon}, 
\authorsc{Raj Barik}, 
\authorsc{Ganesh Bikshandi}, 
\authorsc{Bob Blainey}, 
\authorsc{Bard Bloom}, 
\authorsc{Philippe Charles}, 
\authorsc{Perry Cheng}, 
\authorsc{David Cunningham},
\authorsc{Christopher Donawa}, 
\authorsc{Julian Dolby}, 
\authorsc{Kemal Ebcio\u{g}lu},
\authorsc{Stephen Fink},
\authorsc{Robert Fuhrer},
\authorsc{Patrick Gallop}, 
\authorsc{Christian Grothoff}, 
\authorsc{Hiroshi Horii}, 
\authorsc{Kiyokuni Kawachiya}, 
\authorsc{Allan Kielstra}, 
\authorsc{Sreedhar Kodali}, 
\authorsc{Sriram Krishnamoorthy}, 
\authorsc{Yan Li}, 
\authorsc{Bruce Lucas},
\authorsc{Yuki Makino}, 
\authorsc{Nathaniel Nystrom},
\authorsc{Igor Peshansky}, 
\authorsc{Vivek Sarkar},
\authorsc{Armando Solar-Lezama},  
\authorsc{S. Alexander Spoon}, 
\authorsc{Toshio Suganuma}, 
\authorsc{Sayantan Sur}, 
\authorsc{Toyotaro Suzumura}, 
\authorsc{Christoph von Praun},
\authorsc{Leena Unnikrishnan},
\authorsc{Pradeep Varma}, 
\authorsc{Krishna Nandivada Venkata},
\authorsc{Jan Vitek}, 
\authorsc{Hai Chuan Wang}, 
\authorsc{Tong Wen}, 
\authorsc{Salikh Zakirov}, and
\authorsc{Yoav Zibin}.


For extended discussions and support we would like to thank: 
Gheorghe Almasi,
Robert Blackmore,
Rob O'Callahan, 
Calin Cascaval, 
Norman Cohen, 
Elmootaz Elnozahy, 
John Field,
Kevin Gildea,
Sara Salem Hamouda,
Michihiro Horie,
Arun Iyengar,
Chulho Kim,
Orren Krieger, 
Doug Lea, 
John McCalpin, 
Paul McKenney, 
Hiroki Murata,
Andrew Myers,
Filip Pizlo, 
Ram Rajamony,
R.~K. Shyamasundar, 
V.~T. Rajan, 
Frank Tip,
Mandana Vaziri,
and
Hanhong Xue.


We thank Jonathan Rhees and William Clinger with help in obtaining the
\LaTeX{} style file and macros used in producing the Scheme report,
on which this document is based. We acknowledge the influence of
the $\mbox{\Java}^{\mbox{\textsc{tm}}}$ Language
Specification \cite{jls2}, the Scala language specification
\cite{scala-spec}, and ZPL \cite{zpl}.
%document, as evidenced by the numerous citations in the text.

This document specifies the language corresponding to Version
\integerversion{} of the implementation. The redesign and reimplementation of arrays and rails was  done by Dave Grove and Olivier Tardieu.
 Version 1.7 of the report was co-authored by Nathaniel Nystrom. The design of structs in \Xten{} was led by Olivier Tardieu and Nathaniel Nystrom.

Earlier implementations benefited from significant contributions by
Raj Barik, 
Philippe Charles, 
David Cunningham,
Christopher Donawa, 
Robert Fuhrer,
Christian Grothoff,
Nathaniel Nystrom,  
Igor Peshansky,  
Vijay Saraswat,
Vivek Sarkar, 
Olivier Tardieu,  
Pradeep Varma, 
Krishna Nandivada Venkata, and
Christoph von Praun.
Tong Wen has written many application programs
in \Xten{}. Guojing Cong has helped in the
development of many applications.
The implementation of generics in \Xten{} was influenced by the
implementation of PolyJ~\cite{polyj} by Andrew Myers and Michael Clarkson.
