\chapter{A Potpourri of Types}\label{chp:tsl}

We have already mentioned a few of the basic types that \Xten{} provides in its
standard library.  We will look at some more of them here.  Our discussion will be far
from complete, but it will at least get you started and aware of where to look
as you need more. A complete API for the library can be found in
\href{http://dist.codehaus.org/x10/xdoc/}{http://dist.codehaus.org/x10/xdoc/}.  

The library is very much a work in progress.
For that reason, some of this chapter may become out-of-date reasonably
quickly, mostly, if not entirely, because \Xten's resources will rapidly become
richer. We will try to keep this guide in step as best we can.    

\section{The Numbers Game}\label{sec:tng}
\subsection{The Players}
\Xten{} provides the usual sorts of basic numeric types:
\begin{description}
\item{\bf Byte, Short, Int, Long: }  8-bit, 16-bit, 32-bit, and 64-bit signed integers.
\item{\bf UByte, UShort, UInt, ULong: }  8-bit, 16-bit, 32-bit, and 64-bit {\em unsigned} integers.
\item{\bf Float, Double: } single and double precision IEEE floating point.
\end{description}
It also supplies a double-precision {\tt Complex} type that Fortran and  C++
programmers have come to expect. All of these types are in the package  {\tt
x10.lang}.

In terms of what these types mean and how they get used, \Xten{} is consistent
with what other languages do, so we won't go into a lot of detail here.  In
particular,  implicit conversions will be performed as needed. For the signed
numeric types, the conversions are between a type  and any type to its right:
\begin{verbatim}
    Byte < Short < Int < Long < Float < Double
\end{verbatim}
And similarly, for unsigned types:
\begin{verbatim}
    UByte < UShort < UInt < ULong < Float < Double
\end{verbatim}

Let us be a little more clear about our vocabulary here:
\begin{quote}
{\bf Coercions:} Suppose we declare a variable ``{\tt i}'' to be an {\tt Int}:
{\tt var i: Int;} The expression ``{\tt i as Long}'' is \Xten's syntax for
saying ``convert {\tt i} to be a {\tt Long}.''  In one sense, this conversion
involves no substantive change to {\tt i}'s value as an integer.  On the other
hand, when we realize {\tt i as Long} in memory, or in CPU register, the full 8
byte representation of a long integer will be used.  This sort of
conversion---in form, but not in substance---is usually called a {\em coercion}.

{\bf Casts:}  Another sort of conversion is that from {\tt Int} to {\tt Byte}. 
The value of the expression ``{\tt i as Byte}'' is the result of masking out the
leading 24 bits of {\tt i} and regarding what is left as a {\tt Byte}, meaning
an integer in the range from -128 to 127.  If, for example, {\tt i} is 4095,
which in hex is {\tt 0xfff}, then masking out the leading bits leaves us with
{\tt 0xff}, so {\tt 4095 as Byte == -1 as Byte}. 

A conversion like that from {\tt Int} to {\tt Byte}, or more dramatically, from
{\tt Long} to {\tt Double} that may involve more than a change to the physical
representation of the value is called a {\em cast}.

The example of ``{\tt 4095 as Byte == -1 as Byte}'' shows that casting can be
dangerous. Here, at least, there is nothing subtle going on, so if that wasn't
the programmer's real intention, there is some chance that the problem can be spotted.
One may not always be so lucky. For example, casting {\tt Int} to {\tt Float}
can lose precision for the same reason that {\tt Int} to {\tt Byte} can, because
{\tt Float} only has 23 bits (plus a sign bit) to hold the value.  It uses the
remaining 8 bits in its 32 bit word for the exponent.  So casting
{\tt 23456789 as Float} yields {\tt 2.3456788E7}.  Note carefully the
last digit.  If that last digit had some significance, you are in trouble---and
the problem is going to be {\em very} hard to see.
\end{quote}

Thus, some of the implicit conversions above, like {\tt Int} to {\tt Long} 
are really coercions, while others, like {\tt Long} to {\tt Float}, are
really casts. Where a cast is what is at stake, \Xten{} usually forces you to
be explicit: ``{\tt Int i = someValue(); Byte b = i {\em as Byte};}''.  The
explicit cast ``{\tt as Byte}'' says you are prepared for the consequences of
the conversion.

\subsection{Signed versus unsigned}
This section is largely an aside for Java programmers, because the distinction
between ``signed''  and ``unsigned'' integers does not exist in Java and so may
not be familiar. What is at stake?

From the dawn of computer history, CPUs have supported two kinds of arithmetic
instructions: one set in which one of the bits that make up the integer's value
is used to hold its sign, and a second set in which all integer values are
treated as non-negative.  Having the extra bit as part of the value doubles the
size of the largest positive integer you can express--something that was pretty
important when the normal machine ``word'' was 16 bits wide.  Today, with 64
bit words becoming the norm, the extra bit is not all that big a deal:
$2^{63}-1$ is plenty big---how often will you really need $2^{64}-1$?

When the Java standard was first developed, the designers
felt that, in their experience, exposing both types, as languages like C had done,
had been a source of more errors than the additional
power warranted.  Thus, unsigned values never made it into Java.  Well, they
made it back into \Xten{}. There {\em are} some arguments for having it, aside
from doubling the largest value, not the least:

When you declare something to be ``unsigned'', you are telling the reader of
your code something:  negative values are neither expected nor welcome,
as, for instance, if you are tracking the number of elements in a list.
It is nice to have code that really says what you really mean.

Another argument is that if you want to use a numeric type to represent one of
the 8-bit ISO character types--\eg{} ISO-8859-1, which handles European
languages nearly completely--then {\tt UByte} is a better match than {\tt
Byte}.  Certainly, everyone {\em thinks} of the mapping from numbers to
characters as being a map from the integers 0, \ldots, 255 to characters. Also
with {\tt UByte}, you can test for control characters with one inequality: they
are precisely those whose value as a {\tt UByte} is less than {\tt 0x20}.
Similar considerations apply to multi-byte encodings of characters.

\section{String Theory}

\Xten{} support for character-based applications is limited at the moment, but
with enough implemented to provide a base for applications that are not
fundamentally about manipulating character data.  Internationalization,
in particular, has only very limited support, and regular expressions have
not yet been implemented. What has been implemented will, with a few exceptions,
not surprise anyone coming from C++ or Java. The same can be said for file input
and output: the implementation is limited, get richer, and not surprising.

The basic types for character work are {\tt x10.lang.Char}, {\tt
x10.lang.String}, and  {\tt x10.util.StringBuilder}. Since {\tt StringBuilder}
is in the package {\tt x10.util}, it must be imported explicitly; the other two
need not be.

\subsection{\tt Char} An instance is a 16-bit Unicode character.
\footnote{Strictly speaking, this is true, but in practice, support for
multi-byte character sets in both {\tt Char} and {\tt String} is very limited. 
For the moment, you would be best advised to stay with 7-bit ASCII or 8-bit ISO
character sets.} Like {\tt Byte} and {\tt Int}, {\tt Char} is a struct, but it
is {\em not} a numeric type: there is {\em no} implicit conversion to any
numeric type. The only explicit conversion to a numeric type is the method {\tt
ord} that returns an {\tt Int}:
\begin{verbatim}
   'a'.ord() == 0x61
\end{verbatim}
will evaluate to {\tt true}. You {\em can} cast integer values to characters
\begin{verbatim} 
   i: Int = ...;
   c: Char = i as Char;
\end{verbatim}
does what you expect: produces the {\tt Char} whose ``{\tt ord}'' is {\tt i}.  

Programmers used to avoiding calls to  {\tt isdigit} by writing 
{\tt '0' <=c \&\& c <= '9'} will find that the compiler
complains that you cannot compare integers and characters.  (This will probably be
fixed in an upcoming release of \Xten{}).  For now, when you
are writing \Xten{}, call {\tt isDigit}, {\tt isLowerCase}, {\tt isSpaceChar},
and so on.   All of these utilities are identical in function with their
counterparts that have been around since the birth of C. 
\ignore{The Java versions are
carefully documented at
\href{http://download.oracle.com/javase/6/docs/api/}
{http://download.oracle.com/javase/6/docs/api/}. 
But: unless you are
really deep into internationalization, your intuition will do when it comes to
using {\tt isLowerCase} and the others correctly, so don't feel you need to go
running off to the Web just yet. }

A 16-bit Unicode character can be written as an escape sequence
$\backslash${\tt u{\em nnnn}}, in which each of the {\tt n}'s is a hex digit. 
As usual, the following escape sequences may be used:
\begin{quote}
 {\tt $\backslash$b == $\backslash$u0008}: backspace BS\\              
 {\tt $\backslash$t == $\backslash$u0009}: horizontal tab HT\\       
 {\tt $\backslash$n == $\backslash$u000a}: linefeed LF\\
 {\tt $\backslash$f == $\backslash$u000c}: form feed FF\\
 {\tt $\backslash$r == $\backslash$u000d}: carriage return CR\\
 {\tt $\backslash$" == $\backslash$u0022}: double quote "\\
 {\tt $\backslash$' == $\backslash$u0027}: single quote '\\
 {\tt $\backslash\backslash$ == $\backslash$u005c}: backslash           
\end{quote}
\subsection{\tt String}

A {\tt String} is an {\em immutable} array of {\tt Char}s.  Immutability means
that, for example, once you declare
\begin{verbatim}
    s: String = "hi";
\end{verbatim}
you can, if you wish, access the first character in {\tt s} as {\tt s(0)}, but
you {\em cannot} reassign via ``{\tt s(0) = 'o';}''.

String literals are delimited by ``double quotes,'' as in {\tt "begin,end"}. 
The same escapes apply within {\tt String} literals as for {\tt Char} literals.

Any object, be it from a class or struct, is converted to a {\tt String} by
calling its method {\tt toString()}, and there is always a default implemention
that is provided and does something sensible.  You are free, of course, to
provide your own implementation in the classes and structs you create.  For
example:
\begin{verbatim}
 1 public class Cast {
 2     public static def main(args: Array[String](1)): Void {
 3        Console.OUT.println("I am a "+ new Cast());
 4     }
 5     public def toString() { return "Cast of Thousands"; }
 6 }
\end{verbatim}

In line 3, the effect of the ``{\tt +}'' operation is to concatenate the two
{\tt Strings}.  When the code is run, the console output is:
\begin{verbatim}
%x10 Cast 
I am a Cast of Thousands
\end{verbatim}

There are some basic methods already implemented for {\tt String}s, and over
time more will be added. 
There are four methods for extracting substrings:
\begin{quote}
\begin{description}
\item[{\tt  substring(from:Int)}:] The argument {\tt from} is the index of the first
character to keep.  For example, , if we have {\tt val s = "misguided"}, then 
{\tt s.substring(3)} will be  {\tt "guided"}.
\item[{\tt substring(from:Int, to:Int)}:] The first argument is again the starting
index, and the second, {\tt to}, is the index 1 greater than where you
wish to stop.  Thus if {\tt val s = "misguided"} as before, then
{\tt s.substring(3, length(s)-1)} will be {\tt \"guide"}---and not (as one might
think) {\tt "guided"}.
\item[{\tt trim()}] Remove leading and trailing white-space: {\tt "  abc ".trim()}
is {\tt "abc")}.  In fact, it trims all ASCII control characters as well.
\footnote{ To be absolutely precise:
\begin{quote}
If an empty string, or one whose the first and last characters are represented
by characters whose codes greater than {\tt 0x20}, which is the ASCII code for
the space character, then to the same string is returned.

If the string has no character with code greater than {\tt 0x20}, 
then an empty string is returned.

Otherwise, if {\tt s} is the string, if {\tt start} is the index of the first character in
{\tt s} whose code is greater than{\tt 0x20}, and if {\tt end} is the index of the last
character in {\tt s} whose code is greater than{\tt 0x20}, then the returned string
is {\tt s.substring(start, end+1)}.
\end{quote}
}
\item[{\tt split(s: String)}]  If you need to break up a comma-separated
list, or something like that, use the {\tt split()} method:
\begin{verbatim}
   val s = "a,b, c,  d";
   val parts = s.split(",");
\end{verbatim}
yields an array of four substrings: {\tt "a", "b", " c"} and {\tt "  d"}.
The argument for {\tt split}  in the C++ runtime is currently a {\tt String},
but in the Java runtime, it is a regular expression.   
Eventually, both will accept a regular expression, which will make it
easier to clean up the white-space trailing the commas at the same
time you get rid of the commas.
\end{description}
\end{quote}


There are also four approaches to comparing {\tt String}s. Let {\tt s} and {\tt
t} be two {\tt String}s:
\begin{quote}
{\bf Equality as objects:}  {\tt s == t} evaluates to true if {\tt s} and
{\tt t} name the same {\tt String} object.  {\tt String} is a class, not a
struct, {\em so what} {\tt s} {\em names is a reference to a } {\tt String}, not
a literal string value.

{\tt s != t} is true when {\tt s} and {\tt t} do {\em
not} name the same {\tt String} object.

{\bf Equality as values: }  The expression {\tt s.equals(t)} evaluates to 
true if the current values of {\tt s} and {\tt t} are the same---in other
words, if the two underlying byte arrays are byte-for-byte the same. 
If you don't care about case, use {\tt s.equalsIgnoreCase(t)} instead.

{\bf Infix comparison operators: } The operators {\tt <}, {\tt <=}, {\tt
>=}, and {\tt >} compare two {\tt String} values using lexicographic
(dictionary) order.   For example, {\tt "good"<"goof"} will evaluate to {\tt
true}, since {\tt d} comes before {\tt f}.

{\bf Comparison methods: } The method {\tt compareTo(s:String):Int}
returns a signed integer that is {\em negative} if {\tt s < t}, 0 if
{\tt s.equals(t)}, and {\em positive} if {\tt s > t}.
The method {\tt compare\-To\-Ignore\-Case(s:String):Int} can be used for
comparisons that are {\em not} case-sensitive.  The notion of ``order'' is
the same as for the infix operators.
\end{quote}

Searching {\tt String}s can be done with the methods
\begin{quote}
\begin{description}
\item[{\tt s.indexOf(what,where}, {\tt s.last\-IndexOf(what, where)}:] 
returns the indices of the first and last occurrence of the {\tt String}
or {\tt Char} "{\tt what}" in {\tt s}, if it is present, -1 otherwise.  The
second argument, {\tt where}, is optional.  It is an integer and specifies
the index of position in {\tt s} at which to begin the search.
\item[{\tt s.starts\-With(what)},  {\tt s.endsWith(what)}:]
returns true or false, true meaning that the {\tt String}  "{\tt what}" 
appears at the beginning or end of {\tt s}.
\end{description}
\end{quote}

For those with fond memories of C's {\tt printf} or Java's {\tt format}, \Xten{}
gives you
\begin{quote}
\begin{verbatim}
format(fmt:String, args:Array[Any]):String
\end{verbatim}
\end{quote}
whose first
argument is a format string like those of its C and Java ancestors, and whose
second argument is the array of whatever is to be formatted.\footnote{The type {\tt
Any}, as we discussed in section \ref{ssec:wasacs}, page \pageref{ssec:wasacs},
is the interface all values have in common.}  For example, executing
\begin{quote}
\begin{verbatim}
s = String.format("%s %d %s %10.2f", ["int",32,"dbl",3.2]);
\end{verbatim}
\end{quote}
leaves {\tt s} with the value {\tt "int 32 dbl\ \ \ \ \ \ \ 3.20"}.

Support for the formatting options is limited at the moment, but, as in our
example, most simple cases work as they have done since {\tt printf} was
introduced some forty years ago.

\subsection{\tt StringBuilder} One often has to build up a string whose
length is not known in advance and which isn't short enough that it makes
sense to just build it up as a concatenation {\tt a+b+c+\ldots}
That is the job of {\tt x10.util.StringBuilder}. We remind you that you have to
import this class, like all others in the {\tt x10.util} package.  For
practical purposes, a {\tt StringBuilder} has two methods: {\tt add(a: Any)} and
{\tt result()}.  The method {\tt add} uses its argument's {\tt toString} method
to compute the {\tt String} to append, and {\tt result} simply returns the {\tt
String} you have been accumulating. Here's an example of how one uses a {\tt
StringBuilder}:
\begin{quote}
\begin{verbatim}
val sb = new StringBuilder();
sb.add(2);
sb.add(" and ");
sb.add(3.0);
sb.add(" is "+5.0);
val s = sb.result();
\end{verbatim}
\end{quote}
The value of {\tt s} will be {\tt "2 and 3.0 is 5.0"}.  

\section{Files}\label{sec:files}
Let's start simply, with the program in figure \ref{fig:flc}.  
\begin{figure}[!htbp]
\hrulefill
\begin{verbatim}
 1 import x10.io.File;
 2 
 3 public class FileIO {
 4    public static def main(args: Array[String](1)) {
 5       val inputPath  = args(0);
 6       val outputPath = args(1);
 7       val I = new File(inputPath);
 8       val O = new File(outputPath);
 9       val P = O.printer();
10       for (line in I.lines()) {
11          P.print(line);
12       }
13       P.flush();
14    }
15 }
\end{verbatim}
\caption{A simple line-reader file copy program,
\href{http://dist.codehaus.org/x10/documentation/guide/src/files/FileIO.x10}{files/FileIO.x10}}\label{fig:flc}
\hrulefill
\end{figure}
The fun begins with line 7.
\begin{description}
\item[lines 7 and 8:] 
The basic drill for reading and writing files is to get hold of a {\tt File}
object, and use it to get a stream either inbound or outbound. 
Executing ``{\tt I = new File(in\-put\-Path)}'' converts
a file-system path into an object that represents
the corresponding file, which may or may not yet exist.
This is more in the
spirit of Java than C++:
this step is bypassed in a lot of C++ applications and only the streams are visible.

The reason for having a separate {\tt File} class
is to be able to perform file-level operations.  For example, the call
{\tt I.exists()}  returns {\tt true} when the file constructed in line 7 actually exists,
and {\tt I.isDirectory()} returns {\tt true} when that file not only exists, but is a
directory.  Like the {\tt String} class, {\tt File} is a work in progress.
You will eventually also be able to do operations like making a directory and
renaming a file.  

Most of the time, you will wind up using an instance of {\tt File} as an argument
to a constructor that builds an input or output stream with some desirable
properties like buffering, or handling line enders appropriately in text files.

\item[line 9:] The call {\tt O.printer()} returns an output stream that is an
instance of the class {\tt
Printer}.  This is the class to which the familiar {\tt Console} streams {\tt
OUT} and {\tt ERR} belong, and it is an example of a stream that helps you
write line ends in an operating system-independent way.  
You have already seen the {\tt println} method in action
already.  The {\tt print} method, which we use
here, is the same as {\tt println}, but does not append
the line ender. 
\footnote{ {\tt Printer} also has a {\tt printf} method that is \Xten's take on
the familiar C function, but where the data to be formatted is most
efficiently passed as an {\tt Array}, although if you have 6 or fewer items
to format, you can just string them out, C-style, as extra arguments.}

When the {\tt Printer P} is created, the file to which {\tt P} refers is opened for
writing {\em and is created if need be.}  If the file exists already, its contents are
discarded.
\item[line 10:]  There's no mystery, we hope, about what the loop in lines 10 to
12 does: iterate over the lines in {\tt I}, copying each to {\tt O} via the stream
{\tt P}.  While it is clear that {\tt I.lines()} represents the sequence of lines in
{\tt I}, it is probably not obvious {\em how} it does.  The answer is that it 
implements the interface {\tt x10.lang.Iterator}.
An {\tt Iterator} is always associated with
some underlying collection---in our example, the sequence the lines in {\tt I}.
If the collection contains elements of type {\tt T}, 
the type is {\tt Iterator[T]}, and it has two methods:
\begin{quote}
{\tt {\bf next(): T}} returns the next element, as yet unseen, from the
collection. 

{\tt {\bf hasNext(): Boolean}} returns {\tt true} if a call to {\tt next()} at this
point will succeed.
\end{quote}
Written in terms of {\tt Iterator}'s methods, our {\tt for} loop looks like
\begin{verbatim}
         for(lines = I.lines(); lines.hasNext(); ) {
             P.print(lines.next());
         }
\end{verbatim}
Although there is some chance that, for some {\tt Iterators},
a smart implementation might generate better
code for the {\tt for...in} loop than for the loop written out explicitly
in terms of the iterator,
one should view the {\tt for...in} loop as
a concise, readable way of writing the loop, not as an optimization.
\item[line 13:] We call {\tt flush()} to make sure that, in case the {\tt
Printer} has been buffering our output, all of it gets sent {\em now}.  Do not
count on the output stream being flushed when the program terminates---even when
it terminates normally. Better safe than sorry here.
\end{description}

If you compile {\tt FileIO.x10} and then execute
\begin{quote}
\begin{verbatim}
%x10 FileIO FileIO.x10 out.put
\end{verbatim}
\end{quote}
in the same directory as {\tt FileIO.x10}, you should find that {\tt out.put}
is a copy of {\tt FileIO.x10}.

This example is fine for text files.  Our next examples
are two classes that show how to read and write a binary file.  

Let's look at the writer first, because we need it to generate input for the
reader.  Our goal is to create a file whose contents is a sequence of 
{\tt Doubles}.  We don't want the {\tt String} literals, we really want to store
the numeric values as they appear in memory.  We'll read the {\tt Strings} in
from the keyboard (the ``standard input stream''):
\begin{verbatim}
 1 public class WriteDBL { // see the file files/WriteDBL.x10
 2    public static def main(args: Array[String](1)) {
 3       val I  = new File(args(0));
 4       val W  = new FileWriter(I);
 5       while(true) {
 6         val line = Console.IN.readLine().trim();
 7         if (line.length() == 0) break;
 8         val dbl = Double.parse(line);
 9         W.writeDouble(dbl);
10       }
11       W.close();
12     } 
13  }
\end{verbatim}
We've skipped the imports here---by now, you know what's needed there, and the
working file is available in
\href{http://dist.codehaus.org/x10/documentation/guide/src/files/WriteDBL.x10}{files/WriteDBL.x10}.
\begin{description}
\item[line 4:] We use the class {\tt x10.io.FileWriter} to stream the output. 
It has methods like {\tt writeInt}, {\tt writeFloat}, and so on, that allow you
to write binary data easily out to a file.  In this example, we are going to use
{\tt writeDouble}.

Just as when one creates a new {\tt Printer}, when one creates a new {\tt
FileWriter}, either a new file will be created, or if a file of that
name already exists, all of its contents will be discarded.  In either event,
you start with an empty file.
\item[line 5:] The keyword {\tt while} introduces a loop.  The loop's syntax is
{\tt while({\em boolean\-Expres\-sion}) {\em statement}}.  The expression is
called the {\em loop's test}, and the statement is the {\em loop's body}. 
When execution reaches the test, it is evaluated, and if the value is {\tt
true}, the body is executed. This sequence---check the test, execute the
body---is continued until the test fails: \ie{} evaluates to {\tt false}.
Execution then passes to the code that immediately follows the body.

In our example, the test is {\em always} true, so the loop is 
terminated from within its body, by the {\tt break} statement on line 7.  
\item[lines 6-7:] We read a line of ordinary text from the standard input
stream and trim any white-space characters from the beginning and end of the
line.  If the line is now empty, we break out of the loop.  The statement ``{\tt
break;}'' causes execution to continue at the first statement following the
loop's body.
\item[lines 8-9:] The input line is converted to a {\tt Double}, and that
value is written, bit-for-bit, to the output file.
\item[line 11:] We explicitly close the stream here.  This should guarantee that
any output that has been buffered, but not yet actually written out, does
get written: in other words, it should guarantee that {\tt flush()} is called
before the file's info is discarded.  You can, of course, call {\tt flush}
explicitly at any point where you want make sure your output to that point is
made persistent. This is particularly worth remembering in the early stages of
debugging, where normal completion may not occur.
\end{description}
We compiled our {\tt WriteDBL.x10} and ran it to create a file for the reader to
read:
\begin{verbatim}
% x10 WriteDBL dbl.out
1.25
...
%
\end{verbatim}
If you look at the bytes in {\tt dbl.out}, the first 8 look like {\tt 0x3ff4000000000000}.
Since $1.25 = 5/4$ in hex is 0x$1.4$, this is about what we'd expect: some 
exponent stuff followed by {\tt 0x14}.

Now it's the reader's turn.  
\begin{verbatim}
 1 public class ReadDBL { // see the file files/ReadDBL.x10
 2    public static def main(args: Array[String](1)) {
 3       val inputPath  = args(0);
 4       val I  = new File(inputPath);
 5       val R  = new FileReader(I);
 6       while(true) {
 7         Console.OUT.println(R.readDouble());
 8       }
 9     } 
10  }
\end{verbatim}
The working copy is in
\href{http://dist.codehaus.org/x10/documentation/guide/src/files/ReadDBL.x10}{files/ReadDBL.x10}.
You can see that it is almost line-for-line the same
as the writer---no surprise, really.  The one thing that is different is
figuring out how to terminate the {\tt while} loop in lines 6-8.  The problem is
that {\tt FileReader} does not offer a method for testing when we are at the end
of a stream.  You might think, therefore, that {\tt readDouble} might return a
nonsense value on end-of-stream, so we could test for that.  It does not.  To
see what's going on, look at the console log we got by running the code shown
with input from the file {\tt dbl.out} that we just created:
\begin{verbatim}
%x10 ReadDBL dbl.out
1.2
2.3
3.4
5.678
x10.io.EOFException
   at x10.io.InputStreamReader.read(InputStreamReader.java:73)
   at x10.io.Marshal$LongMarshal.read(Marshal.java:877)
   at x10.io.Marshal$DoubleMarshal.read(Marshal.java:1057)
   at x10.io.Reader.readDouble(Reader.java:164)
   at ReadDBL.main(ReadDBL.java:107)
   ...
\end{verbatim}
We read the file and got just the output we would have expected, and the first
attempt to read beyond the data we provided caused an ``end-of-file exception''.
Which brings us to our next topic: exceptions.
\section{Exceptions}\label{sec:exc}
\subsection{The Basics}
An exception in programming, as in life, is literally an event out of the
ordinary.  Being out of the ordinary, exceptions are a rarity. If they
happened frequently, they would be part of the rule, and not an exception.
In programming, an exception is a way of performing an {\em abnormal return}
from a method call: 
\begin{quote}
An abnormal return is one where the called routine cannot sensibly return a
value to its caller (or simply return to its caller if no return value is at
stake) because some situation, usually something terribly wrong, has arisen
that the caller may not have enough context to handle.  Often, that context is
only available to a method that is many steps earlier in the call chain.  The
problem is to get control efficiently to where it can be exercised effectively,
without unduly penalizing the ``normal'', exception-free paths through the code.
\end{quote}

Instead of each method in the call chain having to look at the
problem and then having to admit that it doesn't know what to do
about it, we force a method that thinks it {\em can} handle it to say so {\em before}
the problem occurs, not {\em after}. The
construct is called a {\tt try} statement.  If you're not familiar with
it, no problem: we'll describe it in detail in a moment.  The important thing
for now is that not every method need take responsibility for every sort of
problem that might happen while it, or something it calls, is executing.

When running {\tt FileIO.main} for example, the attempts to read and write the
files {\em could} fail, but how often does that really happen? Almost never.
If our {\tt main} really wanted to deal with the
error, it could put the code inside a {\tt try} statement that says it is
willing to deal this exception---we'll show the more cautious code in a
moment.

The lingo of the trade is that exceptions are either ``thrown'' or ``raised.'' 
\Xten{} belongs to the ``throw'' camp.  The methods that want to handle an
exception are said to ``catch'' it.
To say that an \Xten{} method ``throws an exception'' means that execution
reaches an \Xten{} {\tt throw} statement, which looks like:
\begin{quote}
{\tt throw new {\em ExceptionSubclass}({\em explanation});}
\end{quote}
For example, if a file open failed because an incorrect path was provided, you
might see
\begin{verbatim}
   throw new x10.io.FileNotFoundException("Bad path '"+path+"'");
\end{verbatim}
What happens when this statement is executed is that an instance of the
class being thrown, {\tt x10.io.FileNotFoundException} in our example,
is constructed.  The class whose name appears in the {\tt throw} must be a
subclass of {\tt x10.lang.Exception}\footnote{Strictly speaking,  you only need
a subclass of the parent class of {\tt Exception}, which is called ``{\tt
Throwable}''.  As a general rule, application code should use or subclass {\tt
Exception} for problems that application code can reasonably be expected to
handle.  {\tt Throwable} should be reserved for \Xten{} and for situations that
you don't expect application code to handle.}.
The newly created instance has a field that holds the argument in the
constructor we called ``{\tt{\em explanation}}'' above.
The catcher can retrieve it by calling the
method {\tt getMessage()}.  The \Xten{} run-time manager will also provide the
instance with the call chain at the point of the ``throw'', which can be
retrieved (as an array of {\tt Strings}, one for each method in the call chain)
by calling the method {\tt getStackTrace()}.  The \Xten{} run-time manager will
then look back up that call chain until a method is found that says that it
wants to handle any exception whose type is a
subclass of the type named in the {\tt throw}.  Control is returned to that
method at an appropriate point---we'll see where in our next example.

Let's rewrite our {\tt ReadDbl.main} so that it handles all its I/O
problems gracefully (see
\href{http://dist.codehaus.org/x10/documentation/guide/src/files/ReadDBL2.x10}{files/ReadDBL2.x10}).
\begin{verbatim}
 1  public class ReadDBL2 {
 2    public static def main(args: Array[String](1)) {
 3       val inputPath  = args(0);
 4       val I  = new File(inputPath);
 5       var r: FileReader = null;
 6       try {
 7          r = new FileReader(I);
 8          while(true) {
 9             Console.OUT.println(r.readDouble());
10          }
11       } catch(eof: x10.io.EOFException) {
12             Console.OUT.println("Done!");
13       } catch(ioe: x10.io.IOException) {
14         Console.ERR.println(ioe);
15       } finally {
16         if (r != null) r.close();
17      }
18    } 
19 }
\end{verbatim}

Here's the line-by-line:
\begin{description}
\item[line 5:]  Our reader {\tt r} is now a {\tt var}, not a {\tt val}, because
we are going to use it when we reach line 16, and we cannot be sure that the
assignment in line 7 will succeed.  Remember, a mistyped input file path will
cause the constructor to throw an exception---an
{\tt x10.io.File\-Not\-Found\-Excep\-tion} in fact, which happens to be a
subclass of {\tt x10.io.IO\-Excep\-tion}.  So the assignment in line 7 may be
aborted.
\item[line 6:] This is the beginning of the {\tt try} statement, the ``try block''.
The brace following the keyword {\tt try} is required.  The catch blocks in
lines 11 and 13 describe the types of exceptions that this {\tt try} statement
promises to handle.  For example, if the assignment in line 7 does {\em not}
complete normally, the exception thrown will be an {\tt
x10.io.IO\-Excep\-tion}, which will cause control to go to line 14, the body of
the catch block that promises to handle {\tt x10.io.IO\-Excep\-tions}.
\item[lines 8-10:] The {\tt while} loop is unchanged.  It goes its merry way
until the end of file exception is thrown, but now, since there is a catcher at
line 11, control will go to line 12 when the exception is thrown.
\item[line 11:] The try block is ended by the closing brace, ``{\tt \}}''. 
What follows is a catch block, which, in English, says: ``if an exception
was thrown in the try block that is an {\tt x10.io.EOFException}, or is an
instance of a subclass of {\tt x10.io.EOFException}, then return control to my
body, namely, the statement bracketed by my braces.  Set the identifier {\tt eof} 
to the object that was thrown.''  At end-of-file, we don't care about the {\tt
eof}'s message or the call-chain traceback, so we simply ignore {\tt eof}. 
We're just done!

All that remains, once the {\tt println} completes, is to be polite and
close up the {\tt FileReader r}.  Because there is a ``finally block'' at line
15, control goes to its body, and the close is done there.  Control then falls
through to whatever statement (if any) comes next.  If there had been no finally block,
then once the body of the catch block had been executed, control would have
passed to the first statement following the {\em last} catch block.
\item[line 13:] Here we have a second catch block, one that is intended to
catch errors.  When you have several catch blocks, they behave like ``{\tt
if\ldots else if \ldots else if \ldots}'' sequences: the first one that can do
the job is the one that the run-time picks: later blocks are ignored.  {\em
This is very important:}
\begin{quote}
If we had put the {\tt IOException} block {\em first}, then, because {\tt
EOF\-Ex\-cep\-tion} is a subclass of {\tt IO\-Ex\-cep\-tion}, the
{\tt IOException} block would
get control on end-of-file as well as whenever an error occurred.  This is
definitely not what we want, and this can be a pernicious error, because it is
not always obvious what classes are subclasses of what other classes.  This
is why, if we had reversed the order of our catch blocks, the compiler
would have signalled it as an error: ``{\tt Unreachable catch block for
EOFException. It is already handled by the catch block for IOException}''.  Try
compiling 
\href{http://dist.codehaus.org/x10/documentation/guide/src/files/ReadDBL3.x10}{files/ReadDBL3.x10}
to see this in action.
\end{quote}
\item[line 14: ] The argument to {\tt println} is just the exception {\tt ioe},
which may seem surprising: why not {\tt ioe.getMessage()}?  The answer is that
since {\tt ioe} is not a {\tt String}, but is in a context here where a {\tt
String} is expected, the compiler replaces {\tt ioe} with {\tt ioe.toString()}. 
For all exceptions, the method {\tt toString()} simply calls {\tt getMessage()},
so {\tt println(ioe)} winds up being the same thing as {\tt
println(ioe.getMessage())}.
\item[line 15: ] The statement following the keyword {\tt finally} will be
executed whether or not the try block terminates without throwing an exception.
The {\tt finally} block will be executed even if an exception was thrown that is
{\em not} one we have a catcher for in this try statement. One of the catch
blocks may be executed before the entering the {\tt finally}, if there is a
catcher for the error that occurred, but no matter what, control will
reach the {\tt finally} before leaving the method.

That is why we had to be careful to provide an initial value for {\tt r}: if the
constructor in line 7 aborted, control would have gone to the second catch block,
and when it completed, control would wind up in the ``finally'' at line 16.
Because we were careful to initialize {\tt r} in line 5, we are on safe ground
accessing its value here, no matter how we got here.  Actually, the compiler
would have complained if {\tt r} had {\em not} been set along some path that
starts at its declaration and reaches its use in line 16.  It is an error in
\Xten{} for code to read an identifier's value before the identifier has been
initialized.
\end{description}

A {\tt try} statement may have a {\tt finally} but no {\tt catch} blocks: no
matter what exception occurs, the {\tt finally}'s statement will get executed. 
The reason is exactly the sort of thing we saw in this example: if you have
acquired some resource and need to release it, putting the release
in a {\tt finally} guarantees that it happens. A {\tt try} statement
need have no {\tt finally} block. In that case, when the body of the try block
completes, control goes to the first statement that follows the last of its
catch blocks.

\section{Functions}
\subsection{Function Types}
We have already seen examples of how one can create functions in \Xten{} and
pass them as arguments.  Indeed, functions are values like any other from the
point of view of assigning them, passing them around as arguments, and so on. 
As a ``type'', though, they form a distinct unit, being neither class, struct,
nor interface.

The basic syntax for function {\em types}---as opposed to
the syntax for specific method or closure definitions---is:
\begin{quote}
{\tt ({\em arg1Type, arg2Type, \ldots}) => {\em returnType}}
\end{quote}
For example, we could declare
\begin{verbatim}
 var doSum: (Array[Double](1)) => Double;
\end{verbatim}
The value of {\tt doSum}, when it is assigned, will be a function that
takes a single argument, of type a singly-indexed array of {\tt Doubles}, and
returns a value of type {\tt Double}.  When it is helpful for people reading the
declaration to have a name for the argument, you can supply one:
\begin{verbatim}
 var doSum: (a: Array[Double](1)) => Double;
\end{verbatim}
Here it is not very useful, but there obviously are a lot of situations where
the variable name and argument types by themselves do not reveal your intent.

The argument list for a function type may be empty, as it was for our random
number generator in the Monte Carlo calculation in \ref{subs:wcdb}.

Getting back to {\tt doSum}, we know that, being a {\tt var}, it can be set (and
reset!) whenever we wish. It would be polite, however, to assign an initial
value in the declaration that, if nothing like what we eventually want, at least
would betray an attempt to use {\tt doSum} before it was properly set.
Because {\tt Double} is a {\tt struct} we cannot set it to {\tt null}.
For {\tt Doubles} the closest thing to {\tt null} is the value ``not a number!'',
{\tt Double.NaN}---not {\tt 0.0}, which is often
a reasonable value after the fact, as well as initially.
\begin{verbatim}
 var doSum: (Array[Double](1)) => Double
          = (a: Array[Double](1)) => Double.NaN;
\end{verbatim}
or to let it throw an exception:
\begin{verbatim}
 var doSum: (Array[Double](1)) => Double
       = (a: Array[Double](1)): Double => {
            val msg = "doSum called before being set.";
            throw new IllegalOperationException(msg);
          };
\end{verbatim}
These two examples show the two patterns for creating function ``values''.
One is exemplified by the right-hand side of the first assignment, in which
the arrow ``{\tt =>}'' is followed by an expression that is the return value of
the function.  The other possibility is to write out the function body in the
usual statement form, bracketed by braces, which is what we did to throw the
exception.  When you use this format for the function body and want to return a
value, you need to use a {\tt return} statement to do so, just as you would in
a method definition.

Notice that in {\em both} of these assignments, a semicolon follows the body of
the function.  In the first case, it looks natural enough, but in the second, where
the semicolon follows the closing brace, you may feel that it looks
a little strange.  Remember, though, that what we have here is an assignment
(to a {\tt var} of its initial value), not a method definition, 
and assignments expect to be terminated with semicolons.  

The bottom line is that the syntax for function literals is the same as that for
function types, except that for a literal, the body follows the arrow rather
than (as for types) the return type. You need not specify
the return type for a function literal explicitly any more than you need do so
for a method definition, unless (as in our second assignment here) the
literal does not return the type of value you really want.

\subsection{A Few Words On Arguments}
We are going
to give a few examples here that lay out \Xten's rules for what happens when you pass
an argument to method and then assign a value to it.  While this discussion should present
few surprises, particularly to Java programmers, we suggest you bear with us and read on,
as the parallels are not exact.

Consider, to begin with, the following attempt to pass some {\tt Int} arguments into a method.
 %%START X10: src/primitives/TryArgs.x10 tryargs
\begin{verbatim}
1  public class TryArgs {
2     public static def tryargs(val a: Int, b: Int, var c: Int) {
3        a = b = 1;  
4        c = 1;       
5     }
6  }
\end{verbatim}
%%END X10:  src/primitives/TryArgs.x10 tryargs
You'll find this code in 
\xlfilename{tryargs}{primitives/TryArgs.x10}
When you compile it, you will get a pair of error messages:
\begin{verbatim}
% x10c  TryArgs.x10
TryArgs.x10:3: Final variable "b" might already have been
               initialized
TryArgs.x10:3: Final variable "a" might already have been
               initialized
\end{verbatim}
The complaint about attempting to reset ``{\tt a}'' after you declared it to be a
{\tt val} is what you would expect: you cannot assign a value to an already initialized
 {\tt val}.
 \footnote{
 The term ``{\tt final}'' comes from Java's way of describing read-only
 data fields.
 }
Why the same complaint should hold for ``{\tt b}'' is less
obvious, but now you know: {\em unless you explicitly say that an argument is a {\tt var}, it is
going to be a {\tt val}.}   Thus, of the three arguments, only {\tt c} may be reset in the
method body.

The next thing to try is modifying a {\tt var} argument in the body of the method, as in 
\href{http://dist.codehaus.org/x10/documentation/guide/src/fcns/TryArgs2.x10}{fcns/TryArgs2.x10}: 
%%START X10: src/primitives/TryArgs2.x10 tryargs2
\begin{verbatim}
 1  public class TryArgs2 {
 2     public static def tryargs(var c: Int) {
 3        c = 1;   
 4        Console.OUT.print("In tryargs, c is "+c);
 5     }
 6     public static def main(args: Array[String](1)): Void {
 7        var ta2: Int = 0;
 8        tryargs(ta2);
 9        Console.OUT.println(".  In main, ta2 is "+ta2);     
10     }
11  }
\end{verbatim}
