\documentclass[a4paper]{article}


\usepackage{fullpage}

\title{Initialization in X10 - Technical Report}

\author{Yoav Zibin \texttt{yoav.zibin@gmail.com}}

\date{}




\input{../commands}




\begin{document}


\maketitle


\lstset{language=java,basicstyle=\ttfamily\small}

\section{Introduction}
This technical report formalizes the hardhat initialization rules in X10
    using \emph{Featherweight X10} (FX10).
Read first the paper ``Object Initialization in X10" to understand the motivation behind the hardhat rules.
    %the terminology (e.g., raw and cooked objects),
%    the initialization rules of X10,
%    and their connection with X10's concurrent and distributed constructs (\finish, \async, and \code{at}).

\begin{proof}
Proof of the main theorem in the paper.
We will break the proof into several lemmas, each proving points (i)--(iv).
\end{proof}


\begin{Lemma}[closed]
  \textbf{(Closed is preserved)}
    For every closed expression~$\grave{\he} \neq \hl$, and a well-typed heap~$H$,
        if $H,\grave{\he} \reducesto H',\acute{\he}$,
        then $\acute{\he}$ is closed.
\end{Lemma}
\begin{proof}
Let's consider all the possible reduction rules in \Ref{Figure}{reduction}.
We will prove by induction on the size of $\he$.

Consider \RULE{R-Async}, where~$\grave{\he} = \async~\hl;\acute{\he}$:
\[\typerule{
}{
  \async~\hl;\acute{\he},H \reduce \acute{\he},H
}
\]
Then, because $\grave{\he}$ is closed, then $\acute{\he}$ is closed.

Consider \RULE{RC-Receiver}, where~$\grave{\he} = \he.\hm(\ol{\he})$:
\[\typerule{
    \he,H \reduce \he',H'
}{
  \he.\hm(\ol{\he}),H \reduce \he'.\hm(\ol{\he}),H'
}
\]
By induction (the size of~$\he$ is smaller than the size of~$\grave{\he}$), we have that~$\he'$ is closed.
Therefore,~$\acute{\he} = \he'.\hm(\ol{\he})$ is closed.

The only remaining two interesting cases are~\RULE{R-New} and~\RULE{R-Invoke}.
Consider~\RULE{R-Invoke} (\RULE{R-New} is handled similarly), where~$\grave{\he} = \hl'.\hm(\ol{\hl})$:
\[
\typerule{
    H(\hl')=\hC(\ldots)
        \gap
    \mbody{}(\hm,\hC)=\ol{\hx}.\he
}{
  \hl'.\hm(\ol{\hl}),H \reduce [\ol{\hl}/\ol{\hx},\hl'/\this]\he,H
}
\]
Note that \he might not be closed, i.e., it may contain method parameters or \this.
However, because the program is well-typed and~$\mbody{}(\hm,\hC)=\ol{\hx}.\he$,
    then \he may only contain either~$\hx_i$ or \this,
    and these are substituted by locations in~$[\ol{\hl}/\ol{\hx},\hl'/\this]\he$,
    thus~$\grave{\he}$ is closed.
\end{proof}

\begin{Lemma}[well-typed]
  \textbf{(Heap is well-typed)}
    For every closed expression~$\grave{\he} \neq \hl$, and a well-typed heap~$H$,
        if $H,\grave{\he} \reducesto H',\acute{\he}$,
        then $H'$ is well-typed.
\end{Lemma}
\begin{proof}
Let's consider all the possible reduction rules in \Ref{Figure}{reduction}.
We will prove by induction on the size of $\he$.

For the following rules we have that~$H=H'$, thus $H'$ is well-typed:
\RULE{R-Finish}
\RULE{R-Async}
\RULE{R-Field-Access}


\RULE{R-Field-Assign}

\end{proof}

%\input{val}

\end{document}
